---
title: "BEE552 Biometry Week11"
author: "Maria Feiler"
date: "04/12/2022"
output: pdf_document
---

```{r setup, include=FALSE}
# Make sure to keep the default for normal processing.
default_output_hook <- knitr::knit_hooks$get("output")

# Output hooks handle normal R console output.
knitr::knit_hooks$set( output = function(x, options) {

  comment <- knitr::opts_current$get("comment")
  if( is.na(comment) ) comment <- ""
  can_null <- grepl( paste0( comment, "\\s*\\[\\d?\\]" ),
                     x, perl = TRUE)
  do_null <- isTRUE( knitr::opts_current$get("null_prefix") )
  if( can_null && do_null ) {
    # By default R print output aligns at the right brace.
    align_index <- regexpr( "\\]", x )[1] - 1
    # Two cases: start or newline
    re <- paste0( "^.{", align_index, "}\\]")
    rep <- comment
    x <- gsub( re, rep,  x )
    re <- paste0( "\\\n.{", align_index, "}\\]")
    rep <- paste0( "\n", comment )
    x <- gsub( re, rep,  x )
  }

  default_output_hook( x, options )

})

knitr::opts_template$set("kill_prefix"=list(comment=NA, null_prefix=TRUE))

knitr::opts_chunk$set(opts.label="kill_prefix")

library(knitr)
library(readxl)
```

## My Learning Journey

*Over the last week, I participated in Biometry in the following ways:*

- I asked / answered **** questions posed in class.

- I asked **** questions in Slack.

- I answered **0** questions posed by other students on Slack.

- I came to Heather's office hours: **Yes**

- I came to Jose's office hours: **No**

- I met with Heather or Jose separately from office hours: **No**

*Anything not falling into one of the above categories?*  

> **No**

*On a scale of 1 (no knowledge) to 10 (complete expert), how would I rate my comfort with R programming after this week?* 

> **7**

*Any topics from last week that you are still confused about?*

\newpage

## Problem Set

*The data for this problem set come from the very recent paper by Cordero et al. (2022) that has been provided. I cleaned up the raw data and have provided that to you in the file BC-measurements.xls.*

```{r data, message = FALSE, warning = FALSE}
bc <- read_xlsx("BC-measurements.xlsx")

# Cleaning column names
dimnames(bc)[[2]] <- c("Site", "BCEst", "Region", "Depth")
```

*In this problem set, you will be examining differences in Black Carbon in different regions of Antarctica.*

*a. Considering only Regions 1 and 2, use a two sample t-test to test for a statistically significant difference in the amount of Black Carbon measured in the snow in each region. Calculate the test statistic (manually – using the formula you learned), state the distribution of the test statistic under the null hypothesis, and calculate the associated p-value. Explicitly check whether the data for these two groups conform to the equal variance assumption and be sure to use the appropriate t-test.*


*b. Why is this an unpaired two-sample t-test? Assuming infinite power to sample when and where you want to, how would you have designed this study to have used a paired two-sample t-test? (This will probably require reading enough of the paper to get the gist of it.)*


*Confirm your results using the R function "t.test”. Set the options so the test matches that calculated in (A).*


*c. Now considering all three Regions in a single "omnibus" test for differences, complete the following ANOVA table by calculating the sum of squares manually (in other words, you can use R to do the summing, but I want you to work out the sums of squares yourself).*

TABLE

*d. What is the null hypothesis for the full ANOVA model? What is the alternative hypothesis? Are the data consistent with the null hypothesis?*

*e. What is the difference between the ANOVA test and a series of pairwise comparisons using a t-test?*

*f. Do these data actually meet the assumptions of ANOVA? If not, what assumption(s) are violated?*

*g. To review GLMs, we will re-fit this data, but with the Black Carbon measurements rounded to the nearest integer. We will examine whether Black Carbon measurements are a function of measurement depth. Fit a Poisson GLM to these data for the following model:*

$$Y_i \sim Pois(\lambda_i)$$
$$log(\lambda_i)=\beta_0+\beta_1 Depth_i$$

*Include in your answer the results of the model and a plot of the data with the best-fitting model curve (and confidence interval) overlaid on top.*