---
title: "BEE552 Biometry Week 1"
author: "Maria Feiler"
date: "1/25/2022"
output:
editor_options:
  markdown:
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## My Learning Journey

Over the last week, I participated in Biometry in the following ways:

- I asked / answered *2* questions posed in class.

- I asked \_\_\_\_ questions in Slack.

- I answered \_\_\_\_\_ questions posed by other students on Slack.

- I came to Heather's office hours: Yes/No

- I came to Jose's office hours: *No*

- I met with Heather or Jose separately from office hours: *No*

Anything not falling into one of the above categories?

- Facilitated zoom working together session over zoom, but no one showed

On a scale of 1 (no knowledge) to 10 (complete expert), how would I rate
my comfort with R programming after this week? *4*

Any topics from last week that you are still confused about?

\newpage

## Problem Set 

```{r sample}
# Vector with M&M colors in order presented
MMcolors <- c("brown", "yellow", "green", "red", "orange", "blue")

# Create vector with sample counts
mybag <- c(5, 1, 2, 2, 1, 5)
names(mybag) <- MMcolors

```

```{r sample statement, echo = FALSE, comment = NA}
cat("My bag's distribution of M&M colors was:", paste(mybag, MMcolors), sep = "\n"
    )
```

### Question 1 
*What is the FORMULA (i.e., an equation) for the probability
of obtaining any given combination of colors? Please define all
variables used.*

The probability of getting $n_1$ brown, $n_2$ yellow, $n_3$ green, 
$n_4$ red, $n_5$ orange, and $n_6$ blue M&Ms from a bag containing n
M&Ms of all available colors is the binomial coefficient times the joint 
probability of getting brown $n_1$ times, yellow $n_2$ times, green $n_3$ 
times, red $n_4$ times, orange $n_5$ times, and blue $n_1$ times. 

$$
P(AnyColors) = 
\frac{n!}{n_{1}!n_{2}!n_{3}!n_{4}!n_{5}!n_{6}!}
p^{n_1}_1p^{n_2}_2p^{n_3}_3p^{n_4}_4p^{n_5}_5p^{n_6}_6
$$
  
### Question 2 

*a. What is the actual PROBABILITY (i.e., a number, not an
equation) of obtaining your combination of colors assuming that all
colors are equally likely?*

$$p_1 + p_2 + p_3 + p_4 + p_5 + p_6 = 1 $$
If all colors are equally likely in the population, then...

$$p_{any} = p_1 = p_2 = p_3 = p_4 = p_5 = p_6 = 1/6 = 0.167 $$
Using the notation from question 1, the probability of getting the 
distribution found in my bag (sample of $n$) in any order from an 
equally distributed population of M&Ms is...

\begin{align*}
P(MyColors) & = 
\frac{n!}{n_{1}!n_{2}!n_{3}!n_{4}!n_{5}!n_{6}!}
p^{n_1}_1p^{n_2}_2p^{n_3}_3p^{n_4}_4p^{n_5}_5p^{n_6}_6\\

& = \frac{n!}{n_{1}!n_{2}!n_{3}!n_{4}!n_{5}!n_{6}!}
p^{n_1+n_2+n_3+n_4+n_5+n_6}_{any}\\

& =\frac{16!}{5!1!2!2!1!5!}(0.167^{5+1+2+2+1+5})\\

\end{align*}

``` {r find probability given sample and population probabilities function}
# Calculate probability from list of occurrences and known population 
# probabilities
find.prob <- function(x = NULL, probs = NULL){
    # Find binomial coefficient
        # Define top value
        top <- factorial(sum(x))
        
        # Calculate the base factorial values
        base <- c()
        for (i in 1:length(x)) {base[i] <- factorial(x[i])}
        
        # Get binomial coefficient
        binom <- top/(prod(base))
    
    # Find probabilities of each option of that number
        prob <- c()
        for (i in 1:length(x)) {prob[i] <- probs[i]^x[i]}
        
    # Solve for probability
        result <- prod(binom, prod(prob))
}
```

```{r equal probability}
# Create vector with probability of all M&M colors given all colors
# are equally likely
equaldist <- c(1/6, 1/6, 1/6, 1/6, 1/6, 1/6)
names(equaldist) <- MMcolors

# Find answer
ans2a <- find.prob(mybag, equaldist)

```

```{r 2a answer, echo = FALSE, comment = NA}
cat("My bag's probability given an equal distribution of colors is ", 
    ans2a, ".", sep = "")
```

*b. What is the PROBABILITY of obtaining your combination of
colors given the actual distribution of colors (according to the
company): 24% blue, 14% brown, 16% green, 19% orange, 13% red, 14%
yellow?*

\begin{align*}
P(AnyColors) & = 
\frac{n!}{n_{1}!n_{2}!n_{3}!n_{4}!n_{5}!n_{6}!}
p^{n_1}_1p^{n_2}_2p^{n_3}_3p^{n_4}_4p^{n_5}_5p^{n_6}_6 \\

& =\frac{16!}{5!1!2!2!1!5!}(0.14^{5}0.14^{1}0.16^{2}0.13^{2}0.19^{2}0.25^{5})

\end{align*}

```{r actual probability}
# Create vector with probability of all M&M colors given actual
# likelihoods
truedist <- c(0.14, 0.14, 0.16, 0.13, 0.19, 0.24)
names(truedist) <- MMcolors

# Find answer
ans2b <- find.prob(mybag, truedist)

```

```{r 2b answer, echo = FALSE, comment = NA}
cat("My bag's probability given the true distribution of colors is ", 
    ans2b, ".", sep = "")
```

*c. What is the PROBABILITY of obtaining your combination of
colors given the actual distribution of colors in your bag of M&Ms?*

\begin{align*}
P(AnyColors) & = 
\frac{n!}{n_{1}!n_{2}!n_{3}!n_{4}!n_{5}!n_{6}!}
p^{n_1}_1p^{n_2}_2p^{n_3}_3p^{n_4}_4p^{n_5}_5p^{n_6}_6 \\

& =\frac{16!}{5!1!2!2!1!5!}(0.3125^{5}0.0.625^{1}0.1250^{2}0.1250^{2}0.0625^{2}0.3125^{5})

\end{align*}

```{r my probability}
# Create vector with sample proportions
mydist <- c()
for (i in 1:length(mybag)) {mydist[i] <- mybag[i]/sum(mybag)}
names(mydist) <- MMcolors

# Find answer
ans2c <- find.prob(mybag, mydist)

```

```{r 2c answer, echo = FALSE, comment = NA}
cat("My bag's probability given my bag's distribution of colors is ", 
    ans2c, ".", sep = "")
``` 

*d. Comparing 2a-2c, what distribution of colors (equal
probability, company-stated probability, sample-based probability) makes
your bag of M&Ms more likely to have occurred.*

```{r compare probabilities}
# Create vector with answers from 2A, 2B, and 2C
ans2 <- c(ans2a, ans2b, ans2c)
names(ans2) <- c("equal distribution", "true distribution", "my distribution")

# Get the list positions in order from greatest to least
order <- sort.list(ans2, decreasing = TRUE)

```

```{r 2d answer, echo = FALSE, comment = NA}
cat("I am most likely to get the distribution from my bag from a population of M&Ms that matches ", "\n", names(ans2[order[1]]), " of colors.",
    sep = "")
```

### Question 3 
*Write a short R script to simulate the combinations of
colors that would have been possible from your bag of M&Ms assuming the
company-stated color distribution (from 2b) and use this script to
calculate the probability of obtaining the combination in your bag.
There are many ways to solve this problem, but I am looking for you to
solve it by sampling hypothetical bags of M&Ms and using that collection
of bags to calculate the probability of obtaining your specific bag of
M&Ms. [Hint 1: Try the R function 'sample'. Hint 2: Make sure to write
your script in a way that makes it possible to also answer Question 4]
(Copy and paste your script below.)*

Assuming there is an infinite population of M&Ms and sampling does not 
appreciably change the population's color distribution, I will be 
sampling with replacement.

```{r my bag similuation}
# Creating pool to sample, setting the colors to be factors and 
# removing labels
plaindist <- unname(truedist)
plainMMcolors <- as.factor(unname(MMcolors))

pool <- data.frame(Proportion = plaindist, Color = plainMMcolors)

# Defining a the number of samples to be taken (5000000), the number
# of candies to be selected based on my original bag, and the number 
# of candy colors
nit <- 5
nsamp <- sum(mybag)
ncolors <- length(MMcolors)

# Creating dataframe for individual sample iterations
sampit <- list()

# Creating dataframe to store color counts of iterations
allsampit <- data.frame(matrix(ncol = ncolors, nrow = nit))

# Take samples of the colors and store in data.frame
for (i in 1:nit) {
    # Get all sample iterations
        sampit[[i]] <- sample(x = pool[,2],       # Sample colors 
                            size = nsamp,         # Number of candies
                            replace = TRUE,       # With replacement
                            prob = pool[,1]       # Define color probs
                            )
        
    # Collect all iterations into counts data frame
        allsampit[i,] <- table(sampit[[i]])
}

# Run test to get order of color counts and add column names
allsampit[1,] <- table(sampit[[1]])
colnames(allsampit) <- dimnames(table(sampit))[[1]]

# Reorder dataframe according to original distribution
allsampit <- allsampit[, MMcolors]

# Determine how many of the samples fit my original bag distribution
samesample <- c()

# Creates list of either 0 (doesn't match) or 1 (matches)
for (i in 1:nit) {
        samesample[i] <- sum(match(paste(allsampit[i,], collapse = ""), 
                                   paste(mybag, collapse = ""),
                                   nomatch = 0))
        } 
```

```{r 3 answer, echo = FALSE, comment = NA}
cat("The probability of obtaining my bag's distribution of colors is ", 
    sum(samesample)/nit, ".",
    sep = "")
```

### Question 4 
*According to your simulation, what was the most likely
combination of colors? What was the probability of getting that
most-likely combination?*

```{r most likely colors}
# Get the color counts from samples into one string
shortsamples <- c()

for (i in 1:nit) {
        shortsamples[i] <- paste(allsampit[i,], collapse = "")
}

# Get the sample distribution that was most common
mostcommon3 <- names(sort(table(shortsamples), decreasing = TRUE)[1:3])

# Test to make sure there are not two that are equally likely
length(unique(mostcommon3)) == 3

# Get most common distribution by breaking apart the string
mostcommon <- mostcommon3[1]
mostcommondist <- strsplit(mostcommon3[1], split = "")

# Creates list of either 0 (doesn't match) or 1 (matches)
samecommonsample <- c()
for (i in 1:nit) {
        samecommonsample[i] <- sum(match(paste(shortsamples[i], collapse = ""), 
                                   paste(mostcommon, collapse = ""),
                                   nomatch = 0))
}

```

```{r 4 answer, echo = FALSE, comment = NA}
cat("The most common distribution of colors from the simulation:", 
    paste(mostcommondist[[1]], MMcolors),
    sep = "\n")

cat("The probability of this distribution is", sum(samecommonsample)/nit)
```

### Question 5
*If each bag of M&Ms had contained 50 M&Ms, would your
combination of colors (whatever they were) be more or less likely to
have occurred and why?*

My combination of colors would have been less likely to have occurred 
if I had sampled 50 from the true population of M&Ms 

```{r}
# Determine what the number of M&Ms of each color would be if the 
# bag had 50 candies and matched my distribution of M&Ms from earlier
bag50 <- c()
for (i in 1:length(mydist)) {bag50[i] <- mydist[i]*50}

# Calculate probability of sampling 50 candies with my distribution
# from the true populatio distribution of M&Ms
ans5 <- find.prob(bag50, truedist)
```

### Question 6
*Lets say you have a bag of 20 M&Ms and find the first 19 are
all blue. What is the probability that the 20th is also blue?*

Assuming that the bag came from the true global population of M&Ms, 
represented by a distribution presented in question 2b, then the likelihood 
of selecting another blue would be the same as selecting blue the 
previous 19 times. This assumes the same as question 4, wherein the 
selection of one candy does not influence the selection of other 
candies and the samples are completely independent and not influencing 
the color distribution of the population significantly. Therefore, the 
probability of selecting blue is `r truedist["blue"]`.

