---
title: "BEE552 Biometry Week 3 Lab"
author: "Maria Feiler"
date: "2/10/2022"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We can ask many things about a statistical distribution:

1. Look at the probability density function: What is the probability of obtaining X (discrete) or a number in the interval (X1,X2) (continuous)?

2. Look at the cumulative probability: What is the probability of obtaining X < X*?

3. Look at the quantiles of the distributions: The inverse of the cumulative distribution - What is X* such that the cumulative probability of obtaining X < X* is the specified quantile? Quantiles can have any size: Quartiles, deciles, percentiles, etc.

4. Look at samples from the distribution: What does the distribution “look like?”

Four basic functions in R for these questions, and combining them with distribution calls means you can ask these questions about specific distributions.

- d = probability density function

- p = cumulative probability

- q = quantiles of the distribution

- r = random numbers generated from the distribution

For example:  
Normal distribution: dnorm, pnorm, qnorm, rnorm  
Log-normal distribution: dlnorm, plnorm, qlnorm, rlnorm  
Poisson: dpois, ppois, qpois, rpois  

```{r}
# Take 100 random draws from the Standard Normal N(0,1)
data <- rnorm(100, mean = 0, sd = 1)
head(data)
```

```{r hist frequency}
# Make histogram
hist(data)
```

```{r hist density}
# Make histogram using density
hist(data, freq = FALSE)
```

```{r hist combined}
hist(rnorm(100, mean = 0, sd = 1), col = "cornflowerblue")
```

```{r rnorm tests}
hist(rnorm(500, mean = 1, sd = 0.5), col = "cornflowerblue")

hist(rnorm(500, mean = 1, sd = 0.05), col = "cornflowerblue")

hist(rnorm(500, mean = 1, sd = 0.0005), breaks = 100, col = "cornflowerblue")
```

```{r density hist}
# What happens if you add the flag “plot=F?”
hist(rnorm(1000, mean = 0, sd = 1), plot = FALSE)
```

```{r pnorm pdf}
quantiles <- seq(-3.5, 3.5, 0.01) # These are the quantiles

density <- dnorm(quantiles, mean = 0, sd = 1) # dnorm gives the pdf for a given quantile

plot(quantiles, density, type = "l", ylab = "Probability density")
```

```{r pnorm cdf}
cumulative <- pnorm(quantiles, mean = 0, sd = 1)
plot(quantiles, cumulative, type = "l", ylab = "Probability")
```

```{r qnorm}
probability <- seq(0, 1, 0.001)
quantiles <- qnorm(probability, mean = 0, sd = 1)
plot(probability, quantiles, type = "l", ylab = "Quantiles")

quantiles2 <- qnorm(probability, mean = 0, sd = 0.2)
plot(probability, quantiles, type = "l", ylab = "Quantiles")
lines(probability, quantiles2, col = "red")
```

```{r discrete distributions}
count<-rpois(500,lambda=3)
table(count)
mean(count)
sd(count)

```

*Checkpoint 1: Does everyone in your group understand the definition of a standard error?*  
Yes

```{r poisson}
par(mfrow = c(3, 1))  #Use ?par to see what this command does - more on this later

sample1 <- rpois(1000, lambda=3)	
sample2 <- rpois(10000, lambda=3)
sample3 <- rpois(100000, lambda=3)

hist(sample1)
hist(sample2)
hist(sample3)

```

```{r poisson stats}
sd(sample1)
sd(sample2)
sd(sample3)

```

```{r poisson standard error 2000}
par(mfrow = c(1, 1))
sample.size <- 1000
means <- c()

for (i in 1:2000) {
        means <- c(means, mean(rpois(sample.size, lambda = 3)))
}

hist(means)

# Get sample standard errors
s.e.1 <- sqrt(var(rpois(sample.size, lambda = 3))/sample.size)
s.e.2 <- sd(means)

s.e.1
s.e.2
```

```{r poisson standard error 10000}
par(mfrow = c(1, 1))
sample.size <- 10000
means <- c()

for (i in 1:10000) {
        means <- c(means, mean(rpois(sample.size, lambda = 3)))
}

hist(means)

# Get sample standard errors
s.e.1 <- sqrt(var(rpois(sample.size, lambda = 3))/sample.size)
s.e.2 <- sd(means)

s.e.1
s.e.2

```

*Checkpoint 2: How does increasing the sample size change the result?*  
It reduced the standard error because you are (in theory) better representing the population with teh larger sample size.

```{r probability mass function p0.5}
par(mfrow=c(3,3))
p=0.5
plot(seq(0,10),dbinom(x=seq(0,10),size=1,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=1",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),1*p,1*p*(1-p)),col="red",lwd=2)
plot(seq(0,10),dbinom(x=seq(0,10),size=2,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=2",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),2*p,2*p*(1-p)),col="red",lwd=2)
plot(seq(0,10),dbinom(x=seq(0,10),size=3,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=3",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),3*p,3*p*(1-p)),col="red",lwd=2)
plot(seq(0,10),dbinom(x=seq(0,10),size=4,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=4",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),4*p,4*p*(1-p)),col="red",lwd=2)
plot(seq(0,10),dbinom(x=seq(0,10),size=5,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=5",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),5*p,5*p*(1-p)),col="red",lwd=2)
plot(seq(0,10),dbinom(x=seq(0,10),size=6,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=6",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),6*p,6*p*(1-p)),col="red",lwd=2)
plot(seq(0,10),dbinom(x=seq(0,10),size=7,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=7",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),7*p,7*p*(1-p)),col="red",lwd=2)
plot(seq(0,10),dbinom(x=seq(0,10),size=8,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=8",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),8*p,8*p*(1-p)),col="red",lwd=2)
plot(seq(0,10),dbinom(x=seq(0,10),size=9,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=9",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),9*p,9*p*(1-p)),col="red",lwd=2)
```

```{r probability mass function p0.1}
par(mfrow=c(3,3))
p=0.1
plot(seq(0,10),dbinom(x=seq(0,10),size=1,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=1",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),1*p,1*p*(1-p)),col="red",lwd=2)
plot(seq(0,10),dbinom(x=seq(0,10),size=2,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=2",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),2*p,2*p*(1-p)),col="red",lwd=2)
plot(seq(0,10),dbinom(x=seq(0,10),size=3,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=3",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),3*p,3*p*(1-p)),col="red",lwd=2)
plot(seq(0,10),dbinom(x=seq(0,10),size=4,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=4",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),4*p,4*p*(1-p)),col="red",lwd=2)
plot(seq(0,10),dbinom(x=seq(0,10),size=5,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=5",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),5*p,5*p*(1-p)),col="red",lwd=2)
plot(seq(0,10),dbinom(x=seq(0,10),size=6,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=6",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),6*p,6*p*(1-p)),col="red",lwd=2)
plot(seq(0,10),dbinom(x=seq(0,10),size=7,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=7",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),7*p,7*p*(1-p)),col="red",lwd=2)
plot(seq(0,10),dbinom(x=seq(0,10),size=8,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=8",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),8*p,8*p*(1-p)),col="red",lwd=2)
plot(seq(0,10),dbinom(x=seq(0,10),size=9,prob=p),typ="h",lwd=5,xlab="# of successes",ylab="",main="n=9",ylim=c(0,0.6))
lines(seq(0,10,0.1),dnorm(seq(0,10,0.1),9*p,9*p*(1-p)),col="red",lwd=2)
```

```{r MASS}
library(MASS) #this loads the library into the workspace
sample.pois<-rpois(1000,lambda=20)
fit<-fitdistr(sample.pois,"gamma",start=list(shape=20,scale=1))
fit

names(fit)
fit$estimate
fit$estimate[1]
fit$estimate[2]
```

```{r MASS gamma plot}
x.vals <-  seq(from = 5, to = 40, by = 1)
hist(sample.pois, freq = FALSE, breaks = x.vals)
lines(x.vals, dgamma(x.vals, 
                     shape = fit$estimate[1],
                     scale = fit$estimate[2]),
                     col = "blue")
```

```{r MASS norm plot}
fit2<-fitdistr(sample.pois,"normal")

hist(sample.pois, freq = FALSE, breaks = x.vals)
lines(x.vals, dnorm(x.vals, 
                     mean = fit2$estimate[1],
                     sd = fit2$estimate[2]),
                     col = "blue")
```

*Checkpoint 3: We see here that the data, which were generated by a Poisson distribution, are fit quite well by the Normal distribution. If this data were research data, what would be one major clue that the Poisson distribution was more appropriate than the Normal distribution?*
We would prefer the Poisson because our data is discrete, so any predictions we'd make based off this distribution would also be discrete (17 seeds vs 17.3 seeds).

```{r fitdistr tests}
qqplot(x = sample.pois, 
       y = rgamma(1000, 
                  shape = fit$estimate[1],
                  scale = fit$estimate[2])
       )

abline(a = 0, b = 1, col = "red", lwd = 2)
```


```{r checkpoint 4}
bern <- matrix(0, nrow = 1000, ncol = 1000)

for (i in 1:1000){
        bern[i,] <- rbinom(n = 1000,
                           size = 1,
                           prob = 0.5)
}

Sn <- rowMeans(bern)

hist(Sn)
```